---

---
title: "DataViz Club"
format: html
editor: visual
---

### Loading the required libraries.

```{r}
library(ggplot2)
library(dplyr)
library(tidyr)
library(corrplot)
library(ggpubr)
library(lubridate)
library(ggthemes)
library(ggrepel)
library(scales)
library(grid)
library(gridExtra)
library(forcats)
library(ggtext)
library(stringr)
library(patchwork)
library(gridExtra)
library(grid)
library(patchwork)
library(cowplot)
```

### Loading the datasets.

```{r}
load('./13_worldcup.Rdata')

matches <- worldcup[[1]]
worldcups <- worldcup[[2]]
```

### Visualizing an overview of the data sets.

The function glimpse is used to easily visualize the features and values of the data sets.

```{r}
glimpse(worldcups)
```

```{r}
glimpse(matches)
```

Visualzing the features' names of the data sets for future reference when creating the graphs

```{r}
colnames(matches)
```

```{r}
colnames(worldcups)
```

### Imputing missing values

For the data set matches, when the feature win_conditions is found empty it means that no conditions applied, therefore we decided to impute the string 'Normal' to those missing values

```{r}
matches$win_conditions[is.na(matches$win_conditions)] <- "Normal"
print(head(matches$win_conditions,20))
```

For the winning and losing team feature if found empty we assume that the game ended with a draw, therefore having no winning or losing teams. We decided to impute missing values in those columns with the string 'Draw'.

```{r}
matches$winning_team[is.na(matches$winning_team)] <- "Draw"
matches$losing_team[is.na(matches$losing_team)] <- "Draw"
```

### Cleaning Features

The string features will be cleaned and processed to ensure a consistent structure and allow for better visualizations.

```{r}
#Trimming blank spaces at the beginning or end of the string
worldcups <- worldcups %>%
  mutate(across(where(is.character), ~ str_trim(.)))

matches <- matches %>%
  mutate(across(where(is.character), ~ str_trim(.)))
```

In the data set matches the feature stage has some incongruities.

```{r}
print(unique(matches$stage))

#Mutating Final Round to Final
matches$stage[matches$stage == "Final Round"] <- "Final"
print(unique(matches$stage))
```

### Feature Engineering

Converting the date feature on the data set matches into date type to better utilize it during data visualization.

```{r}
# Convert date column to Date type
matches$date <- as.Date(matches$date)
```

We decided to differentiate the teams in east, west or neutral based on their alignment during the Cold War.

```{r}
east_countries <- c(
  "Soviet Union", "East Germany", "Czechoslovakia", "Poland", "Hungary", 
  "Bulgaria", "Romania", "North Korea", "Cuba", "China PR", "Yugoslavia")

west_countries <- c(
  "United States", "England", "France", "West Germany", "Italy", "Netherlands", 
  "Belgium", "Spain", "Sweden", "Norway", "Canada", "Australia", "Israel", 
  "South Korea", "Greece", "Portugal", "New Zealand", "Denmark", 
  "Republic of Ireland", "Brazil", "Argentina", "Mexico", "Colombia", 
  "Japan", "South Africa")

# Create new dataset with Cold War alignment
matches_cold_war <- matches %>%
  select(date, home_team, away_team,winning_team,home_score,away_score) %>%
  mutate(
    home_alignment = case_when(
      home_team %in% east_countries ~ "East",
      home_team %in% west_countries ~ "West",
      TRUE ~ "Neutral"
    ),
    away_alignment = case_when(
      away_team %in% east_countries ~ "East",
      away_team %in% west_countries ~ "West",
      TRUE ~ "Neutral"
  )
  )

east_vs_west <- matches_cold_war %>% 
  filter((home_alignment == "East" & away_alignment == "West") |
         (home_alignment == "West" & away_alignment == "East"))
```

### Visualization

#### Graph 1

Graphing the average goals per game over each 15 years period range for countries both in th eeast, west and neutral. Showcasing the top 5 performing teams per each year range.

```{r}
#Combining the home goals and away goals grouped by team and year to calculate the everage goals per match for each team over each 15 year range.

home_goals<-matches%>%
  group_by(year,home_team)%>%
  summarise(goals=sum(home_score), .groups = 'drop')

away_goals<-matches%>%
  group_by(year,away_team)%>%
  summarise(goals=sum(away_score), .groups = 'drop')

average_goals <- full_join(home_goals, away_goals,
                           by = c('year' = 'year', 'home_team' = 'away_team')) %>%
  mutate(
    total_goals = coalesce(goals.x, 0) + coalesce(goals.y, 0),
    matches_played = (!is.na(goals.x)) + (!is.na(goals.y))
  ) %>%
  mutate(avg_goals = total_goals / matches_played) %>%
  select(year, team = home_team, avg_goals) %>%
  filter(!is.na(team))

#Implementing the 15 years range within the data
average_goals_range <- average_goals %>%
  mutate(year_range = paste0((year %/% 15) * 15, "-", ((year %/% 15) * 15) + 14)) %>%
  group_by(year_range, team) %>%
  summarise(avg_goals = mean(avg_goals, na.rm = TRUE), .groups = 'drop')

#Defining each team's alignment to correctly color the data inputs of the graph
average_goals_range <- average_goals_range %>%
  mutate(alignment = case_when(
    team %in% east_countries ~ "East",
    team %in% west_countries ~ "West",
    TRUE ~ "Neutral"
  ))

#Filtering the top 5 performing teams based on avg_goals for each 15 year range.
top_5_teams <- average_goals_range %>%
  group_by(year_range) %>%
  slice_max(order_by = avg_goals, n = 5) %>%
  ungroup()

#Defining colors based on the team alignment. They will be used throughout the visualizations to maintain consistency.
alignment_colors <- c(
  "East" = "#E41A1C",
  "West" = "#377EB8",
  "Neutral" = "#4DAF4A"  
)

p1<-ggplot(average_goals_range, aes(x = team, y = year_range, size = avg_goals)) +
  
  #Adding a horizontal line for the audience to better flow along the same year range.
  geom_hline(aes(yintercept = year_range), color = "gray70", linetype = "dashed") +
  geom_point(aes(fill = alignment, color = alignment), shape = 21,
             stroke = 0.7, alpha = 0.3) +
  
  #Highlighting the top 5 performing teams with a higher alpha for the audience's easier visualization of them
  geom_point(data = top_5_teams, aes(fill = alignment, color = alignment), shape = 21, stroke = 0.7, alpha = 0.5) +
  
  #Labelling the top 5 performing teams as well as adding their average goals per match within the bubble.
  geom_text_repel(data = top_5_teams, aes(label = team), size = 3,
                  nudge_y = 0.5, direction = "y", box.padding = 0.6, 
                  point.padding = 0.3, segment.size = 0.3) +
  geom_segment(data = top_5_teams, aes(x = team, xend = team, y = year_range, yend = year_range),
               color = "black", size = 0.5) +
  geom_text(data = top_5_teams, aes(label = round(avg_goals, 2)), size = 3, color = "white") +
  
  scale_size_continuous(range = c(3, 15)) +
  scale_color_manual(values = alignment_colors) +
  scale_fill_manual(values = alignment_colors) +
  
  theme_minimal() +
  theme(
    text = element_text(family = 'serif'),
    plot.subtitle = element_text(size = 15, face='bold',margin = margin(t = 10)),
    plot.title = element_text(face = 'bold', size = 30),    #Utilizing the top graph's title as the overall title to the entire visualization
    plot.caption = element_text(size=12),
    axis.text.x = element_blank(),
    axis.ticks.x = element_blank(),
    axis.title.x = element_blank(),
    axis.title.y = element_blank(),
    axis.text.y = element_text(size = 10, face = 'bold'),
    strip.text = element_text(size = 12),
    panel.grid.major.y = element_blank(),
    panel.grid.minor = element_blank(),
    panel.grid.major.x = element_blank(),
    legend.position = "left",
    legend.margin = margin(0, 10, 0, 0)
  ) +
  labs(
    title = "Cold War Team Performance",
    subtitle="Average Goals per Match",
    y = "Year Range",
    size = "Avg Goals",
    fill = "Team Alignment",
    color = "Team Alignment" 
  )
print(p1)
```

Graphing a bar chart to showcase the top performing countries in terms of games winned within the Cold War (1954-1990).

```{r}
#Visualizing the number of winning matches by each team as a horizontal bar chart for the top 15 teams, filtered by year between 1954 and 1990
team_wins <- matches %>%
  filter(year >= 1954 & year <= 1990) %>% 
  group_by(winning_team) %>%
  summarise(winning_games = n()) %>%
  filter(winning_team != 'Draw') %>%
  filter(winning_games >= 5) %>%
  mutate(alignment = case_when(
    winning_team %in% east_countries ~ "East",
    winning_team %in% west_countries ~ "West",
    TRUE ~ "Neutral"
  )) %>%
  
  #Filtering top 15 teams based on the number of winned games to better visualize and avoid overcrowding the plot.
  top_n(15, winning_games)

#Calculating the average number of matches won.
avg_winning_games <- mean(team_wins$winning_games)

#Plotting the horizontal bar chart with the average matches won as a vertical line
p2<-ggplot(data = team_wins, aes(x = winning_games, y = reorder(winning_team, winning_games), fill = alignment)) +
  geom_bar(stat = "identity", alpha = 0.5) +
  geom_vline(xintercept = avg_winning_games, linetype = "dashed", color = "black", size = 0.2) +
  
  #Adding the number of won matches for each team next to its corresponding bar for a better understanding of the data and easier comparison among teams.
  geom_text(aes(label = winning_games), hjust = -0.9, size = 4, color = "black", fontface = "italic", family = 'serif') +
  
  #Filling the bars with the color corresponding to the team's alignment
  scale_fill_manual(values = alignment_colors, name = "Country Alignment") +
  
  labs(
    title = 'Winning Matches (1954–1990)',
    subtitle="Comparison of team wins during the Cold War era, colored by political alignment (East, West, Neutral),\n with an average performance benchmark.",
    x = 'Number of Games Won'
  ) +
  
  theme_minimal() +
  theme(
    text = element_text(family = 'serif'),
    plot.subtitle = element_text(size = 12),
    axis.text.x = element_text(size = 10),
    axis.text.y = element_text(size = 10, face='bold'),
    axis.title.x = element_text(size = 12),
    axis.title.y = element_blank(),
    plot.title = element_text(face = "bold", size = 15),
    legend.position = "none", 
    panel.grid = element_blank(),
    caption.position = 'none' 
  )

print(p2)
```

Graphing via a line and area chart the difference in performance between East and West aligned countries during,before, and after the Cold War.

```{r}
teams_long <- matches %>%
  select(year, home_team, home_score, away_team, away_score) %>%
  pivot_longer(
    cols = c(home_team, away_team, home_score, away_score),
    names_to = c("side", ".value"),
    names_pattern = "(home|away)_(.*)"
  )

#Classifying the teams into the east and west block using the previously defined alignment and excluding those with a neutral alignment
teams_long <- teams_long %>%
  mutate(bloc = case_when(
    team %in% east_countries ~ "Eastern Bloc",
    team %in% west_countries ~ "Western Bloc",
    TRUE ~ NA_character_
  ))

teams_long_filtered <- teams_long %>% 
  filter(!is.na(bloc))

#Summarizing the performance (average goals per match) of each block by year
pre_cold_war <- 1947
cold_war_end <- 1991

year_bloc_summary <- teams_long_filtered %>%
  group_by(year, bloc) %>%
  summarise(avg_goals = mean(score, na.rm = TRUE), .groups = "drop") %>% 
  mutate(period = case_when(
              year < pre_cold_war ~ "Pre-Cold War",
              year >= pre_cold_war & year <= cold_war_end ~ "Cold War",
              year > cold_war_end ~ "Post-Cold War"))

#Pivoting to a wide format to be able to compute the difference between West's performance and East's one.
year_wide <- year_bloc_summary %>%
  pivot_wider(names_from = bloc, values_from = avg_goals) %>%
  mutate(diff = `Western Bloc` - `Eastern Bloc`)

#Plotting the graph
#A positive value/area demonstrates a better performance by the West aligned teams compared to the East ones.
#Setting the maximum and minimum values the y-axis can take as the max and min difference previously calculated.
y_max <- max(year_wide$diff, na.rm = TRUE)
y_min <- min(year_wide$diff, na.rm = TRUE)

p3 <- ggplot(year_wide, aes(x = year, y = diff)) +
  geom_line(color = "black", size = 1) +
  geom_point(aes(color = diff > 0), size = 3) +
  geom_area(aes(fill = diff > 0), alpha = 0.5) +
  #Filling the area graph with the respective colors for the teams' alignment
  scale_fill_manual(values = c("TRUE" = "#377EB8", "FALSE" = "#E41A1C"), guide = FALSE) +
  scale_color_manual(values = c("TRUE" = "#377EB8", "FALSE" = "#E41A1C"), guide = FALSE) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  #Implementing vertical lines to distinguish the periods before, during and after Cold war
  geom_vline(xintercept = pre_cold_war, linetype = "dashed", color = "gray40", alpha = 0.7) +
  geom_vline(xintercept = cold_war_end, linetype = "dashed", color = "gray40", alpha = 0.7) +
  #Annotating the period separations
  annotate("text", x = 1930, y = y_max * 0.9, label = "Pre-Cold War", 
           color = "gray30", fontface = "bold", size = 3.5, family='serif') +
  annotate("text", x = 1970, y = y_max * 0.9, label = "Cold War Era", 
           color = "gray30", fontface = "bold", size = 3.5, family='serif') +
  annotate("text", x = 2005, y = y_max * 0.9, label = "Post-Cold War", 
           color = "gray30", fontface = "bold", size = 3.5, family='serif') +
  #Ensuring the x axis matches the period annotations
  scale_x_continuous(limits = c(1930, 2005)) +
  labs(
    title = "Performance Gap Western vs Eastern",
    subtitle = "Positive values indicate better performance by Western teams",
    x = "World Cup Year",
    y = "Difference in Avg Goals per Game"
  ) +
  theme_minimal() +
  theme(
    text=element_text(family='serif'),
    plot.title = element_text(face = "bold", size = 15),
    plot.subtitle = element_text(size = 12),
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    axis.title.x = element_blank()
  )

# Show the plot and the year-wide data
print(p3)
print(year_wide)
```

```{r}
#Initiating a variable to add space between the bottom graphs to avoid their overlap and esnure a better visualization
spacer <- ggplot() + theme_void()

#Creating the bottom row with the combination of the second and third graph with the spacer in between
bottom_row <- plot_grid(
  p2, spacer, p3, 
  ncol = 3, 
  rel_widths = c(1, 0.01, 1),
  align = "h"
)

#Combining the bottom row with the first plot at the top of it
final_plot <- plot_grid(
  p1, bottom_row, 
  ncol = 1, 
  rel_heights = c(1, 0.8)
)

print(final_plot)
```

```{r}
ggsave('final_plot.jpg',plot = final_plot,width = 20, height = 11, dpi = 1000)
```
